---
title: "Importing data in Hierfstat"
author: "Jerome Goudet"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Importing data in Hierfstat}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage{utf8}{inputenc}
---

#Introduction

This vignette documents how to import or enter genotypic data for the `hierfstat` package.  Originally this package was written to estimate and test hierarchical F-statistics, but was then further developped and now include almost all features of the  [Fstat program](http://www2.unil.ch/popgen/softwares/fstat.htm) (no longer maintained), as well as several others. 



#Format required by most functions in *Hierfstat*

The data types that `hierfstat` can analyse are haploid or diploid, unphased, multilocus genotypes. Note that each data set must be made of only one ploidy level.  

The basic data structure required by most `Hierfstat` function is a data frame with the first column containing a population identifier (preferably a number), and the next $nl$ columns the genotype at each of $nl$ loci.  

In `hierfstat`, alleles are encoded as 1, 2 or 3 digits numbers, and genotypes are encoded as numbers with the two alleles collated (as in pasted together).  Other type of data can be imported (see below) but for the time being we focus on the primary data type. Thus imagine that you have an individual genotyped at a microsatellite locus with allele length `120` and `124`, the way to encode it for `hierfstat` is either `120124`or `124120`.   If the data are SNPs, each allele at a locus could be encoded as `1` and `2`, or you may decide to keep the correspondence between nucleotides and alleles (e.g. `1, 2, 3, 4` for `A, C, G, T`). Thus, if the two alleles at a SNP locus are `A` and `T` and an individual is heterozygote, it could be encoded as `14`or `41`.  

Example data sets are included in `hierfstat`. For instance:

```{r}
library(hierfstat) #load the library
data(diploid) # info about this data set with ?diploid
head(diploid)
```

The first individual (first row of the diploid data frame) belongs to population 1. Its genotype at `loc-1` is `44`, thus homozygote for allele `4`.  It is heterozygote for alleles `3`and `4` at both `loc-2` and `loc-3`, and homozygote for allele `3` at `loc-4` and finaly homozygote for allele `4` at `loc-5`.  In fact, `loc-1` and `loc-4` are monomorphic, meaning that only one allele is present in all individuals from all populations.

If a genotype is missing, it is encoded as `NA`. For instance, the fourth individual has not been typed at `loc-3`, nor did the 6th individual for the same locus. 

The first column of this dataframe contains the identifier of the population to which the individual belongs. We can find how many individuals were typed in each population by using teh command table:

```{r}
table(diploid[,1])
```


As another example, we look at dataset `cont.isl99`, a data frame where alleles are encoded as 2 digits numbers:

```{r}
data(contisl99)
head(cont.isl99)
```

the first individual is homozygous for allele `74` at the first locus (`loc.1`) and heterozygous fore alleles `19` and `55` at the second locus.   The genotype could have been written `5519` instead of `1955`, it does not matter.  Note the genotype of the 3rd and fourth individual at the first locus.  They both carry allele `8`, which is in fact encoded as `08`.  When it comes first, the leading 0 disappears, but it must be present in second position. Hence genotype `874`, `0874` and `7408` are the same, but different from genotype `748` who would be understood by hierfstat as an individual heterozygous for alleles `07` and `48`.

Last point: alleles for all loci to be analysed simultaneously must be encoded with the same number of digits.


#Importing data files 

Often the data to be imported are in a text file. If this is the case, the easiest way to import the file into `R` is via one of the workhorse of R, the `read.table` function.

##Importing FSTAT data files

If the data are in the `FSTAT` format, they can be readily imported using the function `read.fstat`:

```{r}
dip<-hierfstat::read.fstat(system.file("extdata","diploid.dat",package="hierfstat"))
head(dip)
```


##Importing from adegenet: genind objects



```{r,message=FALSE }
library(adegenet)
data(nancycats)
head(genind2hierfstat(nancycats)[,1:10]) # only the first 10 loci
#basic.stats(nancycats)
#genet.dist(nancycats)
data(H3N2)
head(genind2hierfstat(H3N2,pop=rep(1,dim(H3N2@tab)[1]))[,1:10]) # only the first 10 positions
#basic.stats(genind2hierfstat(H3N2,pop=rep(1,dim(H3N2@tab)[1])),diploid=FALSE)
data(eHGDP)
head(genind2hierfstat(eHGDP))[,1:11] 
```
 

##importing VCF files 

Variant Call Format (VCF) files have become a standard for genomic data. This is the storage format used for the [1000 genomes](https://www.internationalgenome.org/) for instance. Importing VCF files 

 
##Importing from Quantinemo

[Quantinemo](https://www2.unil.ch/popgen/softwares/quantinemo/) is a genetic simulation program for markers and traits. Data generated by `Quantinemo` can be imported using the function `read.fstat` if the `save_genotype` setting in `quantinemo` is set to $1$.  If the `Quantinemo` output is set to $2$ (extended), 6 extra columns are outputed and these can be read with `hierfstat`using the function  `qn2.read.fstat`. The component  `$dat` of the object return by this function contains the genotypes of the  individuals simulated, while the component `$sex`contains its sex, the component `$ped` the individuals' pedigree and the componnet `$w` their fitness.  For more details about the extended `FSTAT` format of `Quantinemo`, see its manual. 


```{r}
dat<-qn2.read.fstat(system.file("extdata","qn2_sex.dat",package="hierfstat"))
names(dat)
head(dat$sex)
head(dat$dat[,1:10])
#sexbias.test(dat[[1]],sex=dat[[2]])
```


##Importing from ms

The program [ms](http://home.uchicago.edu/rhudson1/source/mksamples.html) of Hudson is commonly used to generate genomic data.   

I briefly discussed the `ms` software.  its output looks like this:

  ms 200 100 -t 20 -I 2 100 100 40 -n 2 0.01 
  
  29161
  
  //
  
  segsites: 23
  
  positions: 0.0689 0.2534 0.3219 0.3350 0.3547 0.3768 0.4339 0.4359 0.4388 0.4694 0.5003 0.5422 0.6575 0.6985 0.7059 0.7147 0.7453 0.7709 0.7891 0.8439 0.8779 0.8857 0.9380 
  
  00100001100000000000000
  
  00001001000000000001000

The first line is the `ms` command line, and it instructed the program to simulate 2 populations, with $\theta=2N_0\mu=20$.  The 2 populations differ in size and the smallest (the second) is a 100th of the first. The two populations exchange $4Nm=40$ migrants per generation.  100 chromosomes are sampled from each population, and this is repeated a 100 times. 

The genetic data itself comes as a serie of 0 and 1, collated one to the other.  These are the SNP sites, with 0 being the ancestral state and 1 the derived state. 

the function `read.ms` allows reading this into R:

```{r}
msdat<-read.ms(system.file("extdata","2pops_asspop.txt",package="hierfstat"),what="SNP")
dim(msdat) # 2nd number is the number of SNPs+1
head(msdat[,1:11]) # first 10 loci  
table(msdat[,1]) # how many inds per pop
```



```{r}
betas(msdat,diploid=FALSE)$betaiovl # population specific FSTs
```